---
title: "Preliminary Results Summary 6-16-17"
author: "Cecina Babich Morrow"
date: "June 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\ 

The following is an overview of what I have been working on.

##Loading packages and creating color defaults
```{r packages_colors, warning=FALSE}
library(readr)
library(lattice)
library(RColorBrewer)
library(ggplot2)
library(tidyr)
library(hypervolume)

myColours <- brewer.pal(6,"Set2")

my.settings <- list(
  superpose.polygon=list(col=myColours[1:4], border="transparent"),
  strip.background=list(col=myColours[6]),
  strip.border=list(col="black")
)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```

##Importing the data

Prior to importing the data, I replaced all of the -999 values with NA in the .csv file.

```{r load_data}
Amniote_Database_Aug_2015 <- read_csv("C:/Users/Cecina/Desktop/Amniote_Database_Aug_2015.csv")
Amniote_Database_Aug_2015$weaning_weight_g<-as.numeric(Amniote_Database_Aug_2015$weaning_weight_g)
```

##Calculating Charnov's invariants

###$C\cdot E$

The first invariant trait is $C\cdot E$, where $C$ is reproductive effort and $E$ is average adult lifespan. Thus $C\cdot E$ represents the fraction of body mass allocated to reproduction per death (Charnov 2002).

Charnov (2002) defines $C$ as $R/m$, where $R$ is the average reproductive allocation per unit time and $m$ is average adult mass.

\ 

####$R$

Charnov defines $R$ as an organism's average allocation to reproduction per unit time (Charnov 2002). To calculate this value, I first multiplied the litter or clutch size of the organism by the number of litters or clutches per year to determine the number of offspring in a year. Then, I multiplied this number by the mass of offspring at independence. The mass at independence was represented by fledging mass for birds, weaning weight for mammals, and hatching size for reptiles. I selected hatching size, rather than egg mass, for the reptiles because the database has better trait coverage for reptile hatching size.


```{r calculate_R, results="hide"}
for (i in 1:nrow(Amniote_Database_Aug_2015)) {
  print(i)
  if(Amniote_Database_Aug_2015$class[i]=='Mammalia') {
    Amniote_Database_Aug_2015[i, 'R'] <- Amniote_Database_Aug_2015[i,"litter_or_clutch_size_n"]*Amniote_Database_Aug_2015[i,"litters_or_clutches_per_y"]*Amniote_Database_Aug_2015[i,"weaning_weight_g"]
  }
  else if(Amniote_Database_Aug_2015$class[i]=='Reptilia') {
    Amniote_Database_Aug_2015[i, 'R'] <- Amniote_Database_Aug_2015[i,"litter_or_clutch_size_n"]*Amniote_Database_Aug_2015[i,"litters_or_clutches_per_y"]*Amniote_Database_Aug_2015[i,"birth_or_hatching_weight_g"]
  }
  else if(Amniote_Database_Aug_2015$class[i]=='Aves') {
    Amniote_Database_Aug_2015[i, 'R'] <- Amniote_Database_Aug_2015[i,"litter_or_clutch_size_n"]*Amniote_Database_Aug_2015[i,"litters_or_clutches_per_y"]*Amniote_Database_Aug_2015[i,"fledging_mass_g"]
  }
}
```

Thus $R$ is the amount of mass allocated by a parent to offspring every year.

```{r R_available}
length(Amniote_Database_Aug_2015$R[!is.na(Amniote_Database_Aug_2015$R)])
length(Amniote_Database_Aug_2015$R[!is.na(Amniote_Database_Aug_2015$R) & Amniote_Database_Aug_2015$class=="Aves"])
length(Amniote_Database_Aug_2015$R[!is.na(Amniote_Database_Aug_2015$R) & Amniote_Database_Aug_2015$class=="Mammalia"])
length(Amniote_Database_Aug_2015$R[!is.na(Amniote_Database_Aug_2015$R) & Amniote_Database_Aug_2015$class=="Reptilia"])

```

There are 1964 species in the database with $R$ values: 197 birds, 942 mammals, and 825 reptiles.

\ 

####$C$

Then I calculated $C=R/m$:

```{r calculate_C}
Amniote_Database_Aug_2015$C<-Amniote_Database_Aug_2015$R/Amniote_Database_Aug_2015$adult_body_mass_g
```

```{r}
length(Amniote_Database_Aug_2015$C[!is.na(Amniote_Database_Aug_2015$C)])
```

There are 1957 species with value for $C$.

\ 

\ 

Finally, I calculated $C\cdot E$:

```{r calculate_C_E}
Amniote_Database_Aug_2015$C_E<-Amniote_Database_Aug_2015$C*Amniote_Database_Aug_2015$longevity_y

length(Amniote_Database_Aug_2015$C_E[!is.na(Amniote_Database_Aug_2015$C_E)])
length(Amniote_Database_Aug_2015$C_E[!is.na(Amniote_Database_Aug_2015$C_E) & Amniote_Database_Aug_2015$class=="Aves"])
length(Amniote_Database_Aug_2015$C_E[!is.na(Amniote_Database_Aug_2015$C_E) & Amniote_Database_Aug_2015$class=="Mammalia"])
length(Amniote_Database_Aug_2015$C_E[!is.na(Amniote_Database_Aug_2015$C_E) & Amniote_Database_Aug_2015$class=="Reptilia"])
```

There are 1416 species with values for $C\cdot E$: 185 birds, 897 mammals, and 334 reptiles.

\ 

###$E/ \alpha$

The second invariant trait is $E/ \alpha$, where $E$ is once again average lifespan and $\alpha$ is the time to female maturity. This ratio represents the trade-off between life-span and time to reproduction.

In order to keep the ratio dimensionless, I converted the longevity values from years to days.

```{r calculate_E_alpha}
Amniote_Database_Aug_2015$E_alpha<-Amniote_Database_Aug_2015$longevity_y*365/Amniote_Database_Aug_2015$female_maturity_d

length(Amniote_Database_Aug_2015$E_alpha[!is.na(Amniote_Database_Aug_2015$E_alpha)])
length(Amniote_Database_Aug_2015$E_alpha[!is.na(Amniote_Database_Aug_2015$E_alpha) & Amniote_Database_Aug_2015$class=="Aves"])
length(Amniote_Database_Aug_2015$E_alpha[!is.na(Amniote_Database_Aug_2015$E_alpha) & Amniote_Database_Aug_2015$class=="Mammalia"])
length(Amniote_Database_Aug_2015$E_alpha[!is.na(Amniote_Database_Aug_2015$E_alpha) & Amniote_Database_Aug_2015$class=="Reptilia"])
```

There are 3126 species with values for $E/ \alpha$: 1095 birds, 1801 mammals, and 230 reptiles.

\ 

###$I / m$

The third invariant trait is $I / m$, where $I$ is the mass of an offspring at independence and $m$ is average adult body mass. This ratio describes the relative size of an independent offspring.

\ 

####$I$

First I calculated $I$ according to the same definitions I used when calculating $R$: fledging mass for birds, weaning weight for mammals, and hatching mass for reptiles.

```{r calculate_I, results="hide"}
for (i in 1:nrow(Amniote_Database_Aug_2015)) {
  print(i)
  if(Amniote_Database_Aug_2015$class[i]=='Mammalia') {
    Amniote_Database_Aug_2015[i, 'I'] <- Amniote_Database_Aug_2015[i,"weaning_weight_g"]
  }
  else if(Amniote_Database_Aug_2015$class[i]=='Reptilia') {
    Amniote_Database_Aug_2015[i, 'I'] <- Amniote_Database_Aug_2015[i,"birth_or_hatching_weight_g"]
  }
  else if(Amniote_Database_Aug_2015$class[i]=='Aves') {
    Amniote_Database_Aug_2015[i, 'I'] <- Amniote_Database_Aug_2015[i,"fledging_mass_g"]
  }
}
```

\ 

\ 


Then I calculated $I/m$:

```{r calculate_I_m}
Amniote_Database_Aug_2015$I_m<-Amniote_Database_Aug_2015$I/Amniote_Database_Aug_2015$adult_body_mass_g

#How many non-NA values for I/m?
length(Amniote_Database_Aug_2015$I_m[!is.na(Amniote_Database_Aug_2015$I_m)])
#How many birds?
length(Amniote_Database_Aug_2015$I_m[!is.na(Amniote_Database_Aug_2015$I_m) & Amniote_Database_Aug_2015$class=="Aves"])
#How many mammals?
length(Amniote_Database_Aug_2015$I_m[!is.na(Amniote_Database_Aug_2015$I_m) & Amniote_Database_Aug_2015$class=="Mammalia"])
#How many reptiles?
length(Amniote_Database_Aug_2015$I_m[!is.na(Amniote_Database_Aug_2015$I_m) & Amniote_Database_Aug_2015$class=="Reptilia"])
```

There are 2127 species with values for $I/m$: 211 birds, 1061 mammals, and 855 reptiles.

\ 

##Trait Coverage

Next I determined how many species had values for all three of the invariant traits.

```{r complete_cases}
#How many species have values for E/alpha, C*E, and I/m?
sum(complete.cases(Amniote_Database_Aug_2015[,c(39,41,42)]))
#How many birds?
length(Amniote_Database_Aug_2015$species[complete.cases(Amniote_Database_Aug_2015[,c(39,41,42)]) & Amniote_Database_Aug_2015$class=="Aves"])
#How many mammals?
length(Amniote_Database_Aug_2015$species[complete.cases(Amniote_Database_Aug_2015[,c(39,41,42)]) & Amniote_Database_Aug_2015$class=="Mammalia"])
#How many reptiles?
length(Amniote_Database_Aug_2015$species[complete.cases(Amniote_Database_Aug_2015[,c(39,41,42)]) & Amniote_Database_Aug_2015$class=="Reptilia"])
```

1138 species have complete cases for all three traits: 172 birds, 849 mammals, and 117 reptiles.

```{r bargraph}
require(tidyr)
traitcoverage<-data.frame(Class=c("Aves","Mammalia","Reptilia"),C_E=numeric(3),E_alpha=numeric(3),I_m=numeric(3),All=numeric(3))
for(i in 1:3){
  traitcoverage$C_E[i]<-length(Amniote_Database_Aug_2015$C_E[!is.na(Amniote_Database_Aug_2015$C_E) & Amniote_Database_Aug_2015$class==traitcoverage$Class[i]])
}
for(i in 1:3){
  traitcoverage$E_alpha[i]<-length(Amniote_Database_Aug_2015$E_alpha[!is.na(Amniote_Database_Aug_2015$E_alpha) & Amniote_Database_Aug_2015$class==traitcoverage$Class[i]])
}
for(i in 1:3){
  traitcoverage$I_m[i]<-length(Amniote_Database_Aug_2015$I_m[!is.na(Amniote_Database_Aug_2015$I_m) & Amniote_Database_Aug_2015$class==traitcoverage$Class[i]])
}
for(i in 1:3){
  traitcoverage$All[i]<-length(Amniote_Database_Aug_2015$species[complete.cases(Amniote_Database_Aug_2015[,c(39,41,42)]) & Amniote_Database_Aug_2015$class==traitcoverage$Class[i]])
}

traitcoverage$C_E<-as.numeric(traitcoverage$C_E)
traitcoverage$E_alpha<-as.numeric(traitcoverage$E_alpha)
traitcoverage$I_m<-as.numeric(traitcoverage$I_m)
traitcoverage$All<-as.numeric(traitcoverage$All)

traitcoverage$Class<-as.character(traitcoverage$Class)
#Add a row of totals
traitcoverage<-rbind(traitcoverage,c("Total",colSums(traitcoverage[,2:5])))
traitcoverage[4,1]<-"Total"

speciesperinvariant<-gather(traitcoverage,key = Class)
colnames(speciesperinvariant)<-c("Class","Trait","Count")
speciesperinvariant$Count<-as.numeric(speciesperinvariant$Count)

barchart(Count~Class,data=speciesperinvariant,groups=Trait,ylab="Number of Species",auto.key=list(space="top",columns=4),par.settings=my.settings)

```


\ 

##Trait Distributions

I created histograms of body mass and the three invariant traits to examine the distributions. Following the total distribution of all species, I plotted separate distributions for birds, mammals, and reptiles.

Body mass:
```{r bodymass_dist}
ggplot(data=Amniote_Database_Aug_2015,aes(x=log(adult_body_mass_g)))+
  geom_histogram(binwidth = 0.1)
ggplot(data=Amniote_Database_Aug_2015[!is.na(Amniote_Database_Aug_2015$adult_body_mass_g),],aes(x=log(adult_body_mass_g),colour=class))+
  geom_freqpoly(binwidth = 0.1)
```

For the three invariant traits, I included vertical dotted lines showing Charnov's estimates for the trait values of mammals and altricial birds (Charnov 2002). These often aren't very close to the values I've calculated, but they appear remarkably accurate for $I/m$.

$C\cdot E$:
```{r C_E_dist}
ggplot(data=Amniote_Database_Aug_2015,aes(x=log(C_E)))+
  geom_histogram(binwidth = 0.1)
ggplot(data=Amniote_Database_Aug_2015[!is.na(Amniote_Database_Aug_2015$C_E),],aes(x=log(C_E),colour=class))+
  geom_freqpoly(binwidth = 0.1)+
  geom_vline(aes(xintercept=log(5)),color=gg_color_hue(3)[1],lty=2)+
  geom_vline(aes(xintercept= log(1.7)),color=gg_color_hue(3)[2],lty=2)
```

$E/ \alpha$:
```{r E_alpha_dist}
ggplot(data=Amniote_Database_Aug_2015,aes(x=log(E_alpha)))+
  geom_histogram(binwidth = 0.1)
ggplot(data=Amniote_Database_Aug_2015[!is.na(Amniote_Database_Aug_2015$E_alpha),],aes(x=log(E_alpha),colour=class))+
  geom_freqpoly(binwidth = 0.1)+
  geom_vline(aes(xintercept=log(3)),color=gg_color_hue(3)[1],lty=2)+
  geom_vline(aes(xintercept= log(1.35)),color=gg_color_hue(3)[2],lty=2)
```

$I/m$:
```{r I_m_dist}
ggplot(data=Amniote_Database_Aug_2015,aes(x=log(I_m)))+
  geom_histogram(binwidth = 0.1)
ggplot(data=Amniote_Database_Aug_2015[!is.na(Amniote_Database_Aug_2015$I_m),],aes(x=log(I_m),colour=class))+
  geom_freqpoly(binwidth = 0.1)+
  geom_vline(aes(xintercept=log(1)),color=gg_color_hue(3)[1],lty=2)+
  geom_vline(aes(xintercept= log(0.3)),color=gg_color_hue(3)[2],lty=2)
```


##Questionable $I/m$ Values

As seen in the distribution of $I/m$ values, some species have $I/m$ values greater than 1, implying that the offspring are larger at independence than the adults. I created a .csv file of just these species. Most are birds with some mammals and one reptile. For the birds, most of the time the fledglings are only slightly larger than the adults, which seems plausible. Some species, however, clearly have incorrect values for either the juvenile or the adult masses. The .csv and an Excel spreadsheet with my notes on this are both available on the GitHub.

```{r I_m_over1}
I_m_over1<-Amniote_Database_Aug_2015[which(log(Amniote_Database_Aug_2015$I_m)>0),c(1,4,5,7,11,23:25,40,41)]
write.csv(I_m_over1,"I_m_over1.csv")
```


##Creating Hypervolumes

I started by subsetting the dataset to include only those species with values for the three invariant traits.

```{r completecase_species}
desiredcolumns<-c(1:7,11,39,41,42)
completecase_species<-Amniote_Database_Aug_2015[complete.cases(Amniote_Database_Aug_2015$adult_body_mass_g,Amniote_Database_Aug_2015$C_E,Amniote_Database_Aug_2015$I_m,Amniote_Database_Aug_2015$E_alpha),desiredcolumns]
```

Next I log-transformed the body mass and invariant trait values. I also Z-transformed these log-transformed values.

```{r log_and_scale, eval=FALSE}
#Log transform
completecase_species$log_bodymass<-log(completecase_species$adult_body_mass_g)
completecase_species$log_C_E<-log(completecase_species$C_E)
completecase_species$log_I_m<-log(completecase_species$I_m)
completecase_species$log_E_alpha<-log(completecase_species$E_alpha)
#Scale the log transformed traits
completecase_species$scale_log_bodymass<-scale(completecase_species$log_bodymass)
completecase_species$scale_log_C_E<-scale(completecase_species$log_C_E)
completecase_species$scale_log_I_m<-scale(completecase_species$log_I_m)
completecase_species$scale_log_E_alpha<-scale(completecase_species$log_E_alpha)
```

```{r, echo=FALSE}
read.csv("completecase_species.csv")
```


I created hypervolumes for each class using both the log-transformed and scaled trait values. These hypervolumes did not differ much based on whether they have been scaled or not, so we will most likely use the log-transformed trait values (without scaling) moving forward.

###Bird Hypervolumes

The bird hypervolume using log-transformed traits:
```{r bird_log_hyp}
completebirds_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Aves",12:15],name = "completebirds_gaussian")
completebirds_gaussian@Volume
plot(completebirds_gaussian,point.dark.factor=1,color=gg_color_hue(3)[1])
```

Using scaled log-transformed traits:
```{r bird_scale_hyp}
completebirds_scale_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Aves",16:19],
                                             name = "completebirds_scale_gaussian")
completebirds_scale_gaussian@Volume
plot(completebirds_scale_gaussian,point.dark.factor=1,color=gg_color_hue(3)[1])
```

###Mammal Hypervolumes

Log-transformed traits:
```{r mammal_log_hyp}
completemammals_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Mammalia",12:15],
                                             name = "completemammals_gaussian")
completemammals_gaussian@Volume
plot(completemammals_gaussian,point.dark.factor=1,color=gg_color_hue(3)[2])
```

Scaled log-transformed traits:
```{r mammal_scale_hyp}
completemammals_scale_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Mammalia",16:19],
                                               name = "completemammals_scale_gaussian")
completemammals_scale_gaussian@Volume
plot(completemammals_scale_gaussian,point.dark.factor=1,color=gg_color_hue(3)[2])
```

###Reptile Hypervolumes

Log-transformed traits:
```{r reptile_log_hyp}
completereptiles_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Reptilia",12:15],
                                               name = "completereptiles_gaussian")
completereptiles_gaussian@Volume
plot(completereptiles_gaussian,point.dark.factor=1,color=gg_color_hue(3)[3])
```


Scaled log-transformed traits:
```{r reptile_scale_hyp}
completereptiles_scale_gaussian<-hypervolume_gaussian(data = completecase_species[completecase_species$class=="Reptilia",16:19],
                                                name = "completereptiles_scale_gaussian")
completereptiles_scale_gaussian@Volume
plot(completereptiles_scale_gaussian,point.dark.factor=1,color=gg_color_hue(3)[3])
```

###All Classes

Finally, I overlaid the hypervolumes for all three classes to visualize differences.

Log-transformed traits:
```{r three_log_hyp}
plot(hypervolume_join(completebirds_gaussian,completemammals_gaussian,completereptiles_gaussian),
     colors = c(gg_color_hue(3)[1],gg_color_hue(3)[2],gg_color_hue(3)[3]))
```

Scaled log-transformed traits:
```{r three_scale_hyp}
plot(hypervolume_join(completebirds_scale_gaussian,completemammals_scale_gaussian,completereptiles_scale_gaussian),
     colors = c(gg_color_hue(3)[1],gg_color_hue(3)[2],gg_color_hue(3)[3]))
```


